---
title: "Data Import"
output: github_document
---

This document will show how to import data.

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(haven)

#you can also bring in library(janitor) but you dont really need to because the clean names is the only function you need from this. You dont wanna load a lot of packages.

```


## Import the FAS litters CSV

```{r}
litters_df = read_csv("./data/FAS_litters.csv")

litters_df = janitor::clean_names(litters_df)
```

janitor is a package you use to clean the data set. 
It will make things lowercase, get rid of spaces and replace w underscores, etc. 


# Look at the data set

```{r}
litters_df

head(litters_df)
tail(litters_df, 10)
```

```{r eval=FALSE}
view(litters_df)
```

# Learning assessment
## Import FAS pups

Use relative paths.

```{r}
pups_df = read_csv("./data/FAS_pups.csv")

pups_df = janitor::clean_names(pups_df)

pups_df
```

Use absolute path

```{r, eval=FALSE}
pups_df = read_csv("/Users/prade/OneDrive/C Fall 2024/Data Science I/homework/data_wrangling_i/data/FAS_pups.csv")
```
if you move this file, this link will no longer work.

## look at read_csv options

```{r}
litters_df=
  read_csv(
    file="data/FAS_litters.csv",
    col_names=FALSE,
    skip=1
  )
```

###what about missing data

```{r}
library(tidyverse)

litters_df =
  read_csv(
    file = "data/FAS_litters.csv",
    na = c("NA", "", ".")
  )

litters_df=janitor::clean_names(litters_df)

pull(litters_df, gd0_weight)
```
using c function to list the stuff that we consider 'missing'

### what if we code `group` are a factor variable?

```{r}
litters_df=
  read_csv(
    file= "data/FAS_litters.csv",
    na=c("NA", "", "."),
    col_types=cols(
      Group= col_factor()
    )
  )
```

## Import an excel file

```{r}
mlb_df= read_excel("data/mlb11.xlsx", sheet="mlb11")
```

for read_excel you need to load realxl
"range=" allows you to specify a range of columns and rows
sheet is also unique to read_excel.

## Import SAS data

```{r}
pulse_df=read_sas("data/public_pulse_data.sas7bdat")

pulse_df
```
for read_sas you need to load the haven package.

## Never use read.csv()

```{r}
litters_df=read.csv("data/FAS_litters.csv")
```

Never do this either:
```{r}
litters_df$L
```




