---
title: "Data Manipulation"
output: github_document
---

This document will show how to import data.

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
```

This document will show how to *manipulate* data.

Import the two datasets that we're going to manipulate

```{r load}
litters_df=
  read_csv("data/FAS_litters.csv", na=c("NA","","."))

litters_df=janitor::clean_names(litters_df)

pups_df=
  read_csv("data/FAS_pups.csv", na=c("NA","", "."))

litters_df=janitor::clean_names(pups_df)

```

## 'select'

Use `select` to select variable

```{r to_see}
select(litters_df, group, litter_number, gd0_weight)
```

```{r range}
select(litters_df, group:gd18_weight)
```

```{r endin}
select(litters_df, -pups_survive)
```

```{r endin_range}
select(litters_df, -(group:gd18_weight))
```

```{r startsw}
select(litters_df, starts_with("gd"))

select(litters_df, contains("pups"))
```

```{r rename}
select(litters_df, GROUP= group)

rename(litters_df, GROUP=group)

select(litters_df, litter_number, gd0_weight, everything())
#this will order something the way you want. and then keep everything else

relocate(litters_df, litter_number, gd0_weight)
#with this you don't have to say keep everything else
```

```{r learning_assesment}
select(pups_df, litter_number, sex, pd_ears)
# select(dataset, var, var, var)
```

## 'filter'
asking if something is true or false and keeping the true

```{r}
filter(litters_df, gd_of_birth == 20)

#double equal is asking "is this var equal to 20 y/n?"
```

```{r}
filter(litters_df, pups_born_alive > 8)

filter(litters_df, pups_born_alive >= 8)
```

```{r}
filter(litters_df, pups_born_alive != 8)
```

```{r}
filter(litters_df, group == "Low8")

filter(litters_df, group %in% c("Low7", "Low8"))
```

```{r}
filter(litters_df, group %in% c("Low7", "Low8"), pups_born_alive == 8)
```

```{r}
drop_na(litters_df)
# drops any row with any missing data at all

drop_na(litters_df, gd0_weight)
# drops rows with missing data for the gd0 weight

```

```{r}
filter(pups_df, sex == 1)

filter(pups_df, pd_walk < 1, sex == 2)

```

##'mutate'

```{r}
mutate(litters_df, wt_gain=gd18_weight - gd0_weight)
#creating variables that are new w manipulation

mutate(litters_df, sq_pups = pups_born_alive^2)

mutate(litters_df, group = str_to_lower(group))
# lowercases all of the group variable; this is overwriting a variable

mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```

##'arrange'

```{r}
arrange(litters_df, gd0_weight)

arrange(litters_df, desc(gd0_weight))

arrange (litters_df, pups_born_alive, gd0_weight)
```

## PIPING!!

you want to do multiple things in a data set:
dont do this:

```{r}
litters_df= read_csv("data/FAS_litters.csv", na=c("NA", "", "."))

litters_df = janitor::clean_names(litters_df)

litters_df_var = select(litters_df, -pups_born_alive)

litters_with_filter = filter(litters_df_var, group == "Con7")

litters_wt_gain = mutate(litters_with_filter, wt_gain = gd18_weight - gd0_weight)

# awful code
# you could also overwrite but it still sucks
```

```{r dothis}
# %>% and |> are the same thing. use ctrl shift m to make it appear

litters_df=
  read_csv("data/FAS_litters.csv", na=c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  select(-pups_born_alive) %>% 
  filter(group == "Con7") %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )
```

Pipe when the thing isnt the first argument 

```{r}
read_csv("data/FAS_litters.csv", na=c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  mutate (wt_gain = gd18_weight - gd0_weight) %>% 
  lm(wt_gain ~ pups_born_alive, data = _)
```

Data export:
```{r}
litters_df=
  read_csv("data/FAS_litters.csv", na=c("NA", "", ".")) %>% 
  janitor::clean_names() %>% 
  select(-pups_born_alive) %>% 
  filter(group == "Con7") %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )

write_csv(litters_df, "data/cleaned_fas_litters.csv")
```

